---
title: "Introduction to Epidemiology and Biostatistics I"
subtitle: "UW Neurology Fellowship"
author: "Marlena Bannick"
date: "6/18/2020"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(beeswarm)
beeswarm_sim <- function(n, mean1, mean2, add_means=FALSE, add_beeswarm=TRUE){
  group_1 <- rnorm(n, mean=mean1)
  group_2 <- rnorm(n, mean=mean2)
  df <- data.frame(
    group=c(rep("group_1", n), rep("group_2", n)),
    mean=c(group_1, group_2)
  )
  if(add_beeswarm){
    beeswarm(mean ~ group, data=df, col=c("purple", "gold"), pch=16)
  }
  if(add_means){
    boxplot(mean ~ group, data=df, add=add_beeswarm, col="#0000ff22")
  }
}
```

## Introduction

**Marlena Bannick**

PhD Student in Biostatistics at the University of Washington

Math Sciences Researcher at the Institute for Health Metrics and Evaluation

## Goal

Introduce you to epidemiological thinking and key (bio)statistical concepts that you can use to critically interpret scientific studies in health and medicine.

## Learning Objectives I.

- Identify key elements of an epidemiological study and how they relate to the scientific question
- Recognize the basic types of epidemiological study design and identify when each design is appropriate for the scientific question
- Recognize sources of bias in study designs or measurements and understand how they might affect your ability to answer the scientific question

## Learning Objectives II.

- Understand how you can formulate your understanding about a data generating process, assumptions, and a hypothesis to test in a statistical model
- Recognize the distinction between an effect size, a confidence interval, and a p-value as they relate to parameters that are estimated in a statistical model

## Simple Means

```{r}
set.seed(0)
beeswarm_sim(n=10, mean1=2.0, mean2=2.5)
```

## Simple Means

```{r}
set.seed(0)
beeswarm_sim(n=10, mean1=2.0, mean2=2.5, add_means=TRUE)
```

## Small Effect Size, Small Sample Size

```{r}
set.seed(0)
par(mfrow=c(2, 2))
for(i in 1:4){
  beeswarm_sim(n=10, mean1=2.0, mean2=2.5, add_means=TRUE)
}
```

## Large Effect Size, Small Sample Size

```{r}
set.seed(0)
par(mfrow=c(2, 2))
for(i in 1:4){
  beeswarm_sim(n=10, mean1=2., mean2=10., add_means=TRUE)
}
```

## Small Effect Size, Large Sample Size

```{r}
set.seed(0)
par(mfrow=c(2, 2))
for(i in 1:4){
  beeswarm_sim(n=100000, mean1=2.0, mean2=2.5, add_means=TRUE, add_beeswarm=FALSE)
}
```

## Slide with Plot

